---
title: "Exploring rossmann store sales"
author: "NighTurs"
date: "November 5, 2015"
output: html_document
---

The website for this competition is https://www.kaggle.com/c/rossmann-store-sales

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
library(plyr)
set.seed(294)
```

Read data

```{r warning=FALSE, message=FALSE}
train <- fread("data/train.csv", head = T, sep = ',')
stores <- fread("data/store.csv", head = T, sep = ',')
data <- merge(train, stores, by = "Store")
row_count <- nrow(data)
col_count <- ncol(data)
cat("Row count : ", row_count, "; Columns count : ", col_count)
```

First glance at columns and values

```{r warning=FALSE, message=FALSE}
str(data)
```

A bit of transformations, parse dates, convert to factors
```{r warning=FALSE, message=FALSE}
data$Open = factor(data$Open, levels = c(0, 1), labels = c("Closed", "Open"))
data$Promo = factor(data$Promo, levels = c(0, 1), labels = c("None", "Promo"))
data$StateHoliday = factor(data$StateHoliday, levels = c("0", "a", "b", "c"), labels = c("None", "Public", "Easter", "Christmas"))
data$SchoolHoliday = factor(data$SchoolHoliday, levels = c("0", "1"), labels = c("None", "Holiday"))
data$StoreType = factor(data$StoreType, levels = c("c", "a", "d", "b"), labels = c("C", "A", "D", "B"))
data$Assortment = factor(data$Assortment, levels = c("a", "c", "b"), labels = c("Basic", "Extended", "Extra"))
data$Promo2 = factor(data$Promo2, levels = c(0, 1), labels = c("None", "Promo"))
data$PromoInterval = factor(data$PromoInterval, levels = c("", "Jan,Apr,Jul,Oct", "Feb,May,Aug,Nov", "Mar,Jun,Sept,Dec"), labels = c("None", "Jan,Apr,Jul,Oct", "Feb,May,Aug,Nov", "Mar,Jun,Sept,Dec"))
data$Promo2SinceWeek[is.na(data$Promo2SinceWeek)] <- -1
data$Promo2SinceYear[is.na(data$Promo2SinceYear)] <- -1
data$CompetitionOpenSinceMonth[is.na(data$CompetitionOpenSinceMonth)] <- -1
data$CompetitionOpenSinceYear[is.na(data$CompetitionOpenSinceYear)] <- -1
data$CompetitionDistance[is.na(data$CompetitionDistance)] <- -1
data$Date = as.Date(data$Date)
data$DayOfYear = as.integer(format(data$Date, "%j"))
data$Year = as.integer(format(data$Date, "%Y"))
```

Are there still any NA's?
```{r warning=FALSE, message=FALSE}
length(which(is.na(data)))
```

Looking at Sales summaries
```{r warning=FALSE, message=FALSE}
summary(data$Sales)
summary(data[data$Open == "Closed", ]$Sales)
```

Lets remove all rows with store being closed, and also cap Sales by 20000
```{r warning=FALSE, message=FALSE}
data <- data[data$Open == "Open" & data$Sales <= 20000]
data <- data[data$Open == "Open"]
```

Plot sales
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
ggplot(data) + aes(Sales) + geom_histogram(binwidth = 50, fill = "blue", alpha = 0.5)
```

Plot how promo affects sales
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
ggplot(data) + aes(Promo, fill = Promo) + geom_bar()
ggplot(data) + aes(Sales, fill = Promo) + geom_histogram(binwidth = 50, position = "identity", alpha = 0.5)
```

Store types 
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
ggplot(data) + aes(StoreType, fill = Assortment) + geom_bar()

ggplot(data) + aes(StoreType, Sales, fill = Assortment) + geom_boxplot(outlier.size = 1, outlier.colour = "blue")
```

Holidays
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
table(data$StateHoliday)
ggplot(data) + aes(StateHoliday, Sales, fill = StateHoliday) + geom_boxplot(outlier.size = 1, outlier.colour = "blue")

ggplot(data) + aes(SchoolHoliday, fill = SchoolHoliday) + geom_bar()
ggplot(data) + aes(SchoolHoliday, Sales, fill = SchoolHoliday) + geom_boxplot( outlier.size = 1, outlier.colour = "blue")
```

How school holidays are distributed throught weekdays?
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
ggplot(data) + aes(as.factor(DayOfWeek), fill = SchoolHoliday) + geom_bar()
```

Seasonality check
```{r warning=FALSE, message=FALSE,  fig.width = 10, fig.height = 5}
ggplot(data) + aes(DayOfYear, Sales, group = Year, colour = as.factor(Year)) + stat_summary(fun.y = median, geom = "line")

ggplot(data) + aes(as.factor(DayOfWeek)) + geom_bar()

ggplot(data) + aes(as.factor(DayOfWeek), Sales) + geom_boxplot( outlier.size = 1, outlier.colour = "blue")

```
